Ab Initio Effective One-Electron Potentials Revisited: General Theory and Example Applications for Charge-Transfer Energy Calculations

### 2019-05-21 papers/2_oep_technique/paper-r0.tex
Generalized density fitting in incomplete basis sketched

The above technique can be now applied to the CI expansion of the interaction\hyp{}induced property
under the localized basis set approximation 
from Eq.~\eqref{e:delta-p-part}.
%Coulomb\hyp{}like and overlap\hyp{}like 
%terms from Eq.~\eqref{e:delta-p-part-2el}. 
For example, the Coulomb\hyp{}like contributions
can be rewritten without making any further approximation as
%
\begin{equation}
 \sum_{ijkl} 
	{s} \left[ P^{AABB}_{ijkl} \BraKet{AA}{BB} \right]
 \equiv
	\sum_{kl\in B} 
	\tBraKet{k^B}{\hat{v}^{{\rm eff},A}}{l^B}
\end{equation}
%
with
%
\begin{equation}
 \hat{v}^{{\rm eff},A} 
	\equiv \sum_{ij\in A} {s} \left[ P^{AABB}_{ijkl} \hat{v}^{AA}_{ij} \right]  \;.
\end{equation}
%
Similarly for the overlap\hyp{}like term we have
%
\begin{equation}
 \sum_{ijkl} 
	{s} \left[ P^{AAAB}_{ijkl} \BraKet{AA}{AB} \right]
 \equiv
	\sum_{k\in A} \sum_{l\in B} 
	\tBraKet{k^A}{\hat{v}^{{\rm eff},A}}{l^B}
\end{equation}
%
and the associated effective potential reads
%
\begin{equation}
 \hat{v}^{{\rm eff},A} 
	\equiv \sum_{ij\in A} {s} \left[ P^{AAAB}_{ijkl} \hat{v}^{AA}_{ij} \right] \;.
\end{equation}
%
Note that the exchange\hyp{}like terms cannot be represented in terms of OEP's.
Gathering the above arguments, $\Delta^{AB}P$ adopts the following form
%
\begin{multline}
 \Delta^{AB}P \approx \Delta_{\rm ex}^{AB} P
	+ \sum_{i\in A} \sum_{j\in B} 
	  \tBraKet{i}{ \left\{ \hat{v}^{{\rm eff},AB}_{1(ij)} + \hat{v}^{{\rm eff},BA}_{1(ij)}  \right\} }{j}
	 \\ +
	\sum_{i\in A} \sum_{k\in A}
	  \tBraKet{i}{\hat{v}^{{\rm eff},BA}_{2(ik)} }{k}
	 +
        \sum_{j\in B} \sum_{l\in B}
          \tBraKet{j}{\hat{v}^{{\rm eff},AB}_{2(jl)} }{l}
\end{multline}
%
where the exchange\hyp{}like component is
%
\begin{equation}
	\Delta_{\rm ex}^{AB} P = 
	s \sum_{ik\in A} \sum_{jl\in B}
	P^{ABAB}_{ijkl} \BraKet{ij}{kl} \;,
\end{equation}
%
and
the overlap\hyp{}like and the Coulomb\hyp{}like 
OEP's are defined as follows
%
\begin{align}
	\hat{v}^{{\rm eff},BA}_{1(ij)} &\equiv P_{ij}^{AB} \hat{h}
	+ s \sum_{mk\in A}  P^{AAAB}_{mkij} \hat{v}_{mk}^{AA} \\
	%
	\hat{v}^{{\rm eff},BA}_{2(ik)} &\equiv P_{ik}^{AA} \hat{h} 
        + s \sum_{mk'\in A} P^{AAAA}_{ikmk'} \hat{v}_{mk'}^{AA}  \nonumber
        \\
	&+ \frac{1}{2}
        s \sum_{jl'\in B}  P^{AABB}_{ik'jl} \hat{v}_{jl'}^{BB} 
\end{align}
%
and accordingly for the twin operators $\hat{v}^{{\rm eff},AB}_{\text{$1$ and $2$}}$.

@@

An arbitrary one-electron potential of molecule $A$ acting on any state vector 
associated with molecule $A$ can be expanded in an auxiliary space centered 
on $A$ as
\begin{equation}
   \hat{v}^{A}\Ket{i} = \sum_{\xi\eta} \hat{v}^{A}\Ket{\xi} [{\bf S}^{-1}]_{\xi\eta} \BraKet{\eta}{i}
\end{equation}
under the necessary assumption that the auxiliary basis set is nearly complete,
i.e., 
$\sum_{\xi\eta} \Ket{\xi}[{\bf S}^{-1}]_{\xi\eta} \Bra{\eta} \cong 1$.
%In a special case when the basis set is orthogonal (e.g., molecular orbitals)
%the above relation simplifies to
%\begin{equation}
%   v\vert i) = \sum_{\xi} v\vert \xi) ( \xi \vert i)
%\end{equation}
The above general and exact expansion can be obtained by 
utilizing the resolution of identity or, equivalently, density fitting
in the complete space,
%
\begin{equation}
 \hat{v}^{A}\Ket{i} = \sum_{\xi} V^A_{i\xi} \Ket{\xi} \;.
\end{equation}
%
In the above equation,
the matrix ${\bf V}^A$
is the representation of the OEP operator $\hat{v}^{A}$
in the complete basis $\{ \varphi_\xi \}$.
By constructing the least\hyp{}squares objective function 
%
\begin{equation}
 Z[{\bf V}^A] = \int 
                     \left[    v({\bf r}_1) \phi_i({\bf r}_1) - \sum_\xi V^A_{i\xi} \varphi_\xi({\bf r}_1) \right]^2
d{\bf r}_1
\end{equation}
%
and requiring that
%
\begin{equation}
 \frac{\partial Z[{\bf V}^A]}{\partial V^A_{i\mu}} = 0 \text{ for all $\mu$}
\end{equation}
%
one finds the coefficients of the $i$th row of ${\bf V}^A$ to be
%
\begin{equation}
% {\bf v}_i = {\bf a}^{(i)} \cdot {\bf S}^{-1}
  V^A_{i\xi} = \sum_\eta [{\bf S}^{-1}]_{\xi\eta} a^{(i)}_\eta \;,
\end{equation}
%
where
%
\begin{align}
 a^{(i)}_\eta &= \int \varphi^*_\eta({\bf r}) \hat{v}^A \phi_i({\bf r}) d{\bf r} \;,\\  
 S_{\eta\xi}  &= \int \varphi^*_\eta({\bf r}) \varphi_\xi({\bf r}) d{\bf r} \;.
\end{align}
%
%Since matrix elements of an OEP operator in auxiliary space can be computed 
%in the same way as the matrix elements with any other basis function, 
%one can formally write the following identity
%\begin{equation}
% (X \vert v\vert i) = \sum_{\xi\eta} S_{X \xi} [{\bf S}^{-1}]_{\xi\eta} (\eta\vert v\vert i)
%\end{equation}
%where $ X $ is an arbitrary orbital.
%When the other orbital
%does not belong to molecule $A$ but to the (changing) environment, it is 
%straightforward to compute the resulting matrix element, which is simply given as  
%\begin{equation}
%   (j_{\in B} \vert v^A \vert i_{\in A}) = \sum_\xi {S_{j\xi}} {G_{i\xi}}
%\end{equation}
%where $j$ denotes the other (environmental) basis function.
%
The working form of $a^{(i)}_\eta$ can be found by applying 
potential for from Eq.~\eqref{e:v-eff}
along with the spectral representation of the effective density from Eq.~\eqref{e:d-spectral} 
which gives
%
\begin{equation}
 a^{(i)}_\eta = \sum_{x\in A} W_{\eta i}^{(x)} + 
 \sum_{\alpha\beta} D^A_{\alpha\beta} 
  \BraKet{\alpha\beta}{\eta i} \;.
%  \iint \frac{ \phi_\alpha^*({\bf r}_1)  \phi_\beta({\bf r}_1) \varphi^*_\eta({\bf r}_2) \phi_i({\bf r}_2)}
% {\vert {\bf r}_1 - {\bf r}_2 \vert }
% d{\bf r}_1 d{\bf r}_2
\end{equation}
%

@@

Density fitting scheme from previous section has practical disadvantage of a nearly-complete basis set
being usually very large (spanned by large amount of basis set vectors). Any non-complete basis set
won't work in the previous example. Since most of basis sets used in quantum chemistry do not form a nealy complete
set, it is beneficial to design a modified scheme in which it is possible to obtain the effective 
matrix elements of the OEP operator in a incomplete auxiliary space. This can be achieved by minimizing 
the following objective function
\begin{multline}
	Z[{\bf V}^A] = \iint 
        \Bigg\{
                     \left[    v^A({\bf r}_1) \phi_i({\bf r}_1) - \sum_\xi V^A_{i\xi} \varphi_\xi({\bf r}_1) \right]
                  \\ \times \frac{1}{\vert {\bf r}_1 - {\bf r}_2 \vert}  
                     \left[    v^A({\bf r}_2) \phi_i({\bf r}_2) - \sum_\xi V^A_{i\xi} \varphi_\eta({\bf r}_1) \right]
		    \Bigg\}
         d{\bf r}_1 d{\bf r}_2  \;.
\end{multline}
Requesting that 
\begin{equation}
 \frac{\partial Z[{\bf V}^A]}{\partial V^A_{i\mu}} = 0 \text{ for all $\mu$}
\end{equation}
one obtains
\begin{equation}
% {\bf G}^{(i)} = {\bf b}^{(i)} \cdot {\bf A}^{-1}
  V^A_{i\xi} = \sum_\eta [{\bf A}^{-1}]_{\xi\eta} b^{(i)}_\eta \;,
\end{equation}
where 
\begin{align}
 b^{(i)}_\eta &= \iint 
                       \frac{ \varphi^*_\eta({\bf r}_1) \hat{v} \phi_i({\bf r}_2) } 
                            {\vert {\bf r}_1 - {\bf r}_2\vert}  
                 d{\bf r}_1 d{\bf r}_2 \;, \\
 A_{\eta\xi}  &= \iint 
                       \frac{ \varphi^*_\eta({\bf r}_1) \varphi_\xi({\bf r}_2) } 
                            {\vert {\bf r}_1 - {\bf r}_2\vert}  
                 d{\bf r}_1 d{\bf r}_2 \;.
\end{align}
%
%The symbol $ \vert\vert $ is to denote the operator $ r_{12}^{-1}$ and double integration over $ {\bf r}_1 $
%and $ {\bf r}_2 $. 
Note that, while $A_{\eta\xi}$ is a typical 2\hyp{}center ERI,
$b^{(i)}_\eta$ matrix elements are not at all trivial to evaluate
because the OEP operator, which contains integration over an electron coordinate,
is present inside the double integral. Therefore, the following triple integral
has to be computed
%
\begin{equation}
 b^{(i)}_\eta = \iiint 
           \frac{ \varphi^*_\eta({\bf r}_1) \phi_i({\bf r}_2)  \rho^{\rm eff}({\bf r}_3) }
            {\vert {\bf r}_1 - {\bf r}_2 \vert \vert {\bf r}_3 - {\bf r}_2 \vert}
           d{\bf r}_1 d{\bf r}_2 d{\bf r}_3
\end{equation}
%
Computation of all the necessery integrals of this kind is very costly 
and impractical even for medium sized molecules. 
However, one can introduce the effective potential in order to eliminate one integration. 
This can be achieved by performing additional density fitting in a nearly complete intermediate basis 
%
\begin{equation}
 \hat{v}^{\rm eff} \Ket{i} = \sum_\varepsilon^{\rm Int} H_{i\varepsilon} \Ket{\varepsilon}
\end{equation}
%
where the symbol `Int' denotes the intermediate basis from a chosen resolution of identity.

%Therefore, in order to use this generalized density fitting scheme
%one must to compute two\hyp{}centre electron repulsion integrals
%as well as four\hyp{}centre asymmetric electron repulsion integrals of the type $ (\alpha\beta\gamma||\eta) $.
%The evaluation of such integrals is discussed in Appendix~\ref{a:mcmurchie-davidson}.

>>>>

The above technique can be now applied to the CI expansion of the interaction\hyp{}induced property
under the localized basis set approximation 
from Eq.~\eqref{e:delta-p-part}.
%Coulomb\hyp{}like and overlap\hyp{}like 
%terms from Eq.~\eqref{e:delta-p-part-2el}. 
For example, the Coulomb\hyp{}like contributions
can be rewritten without making any further approximation as
%
\begin{equation}
 \sum_{ijkl} 
	{s} \left[ P^{AABB}_{ijkl} \BraKet{AA}{BB} \right]
 \equiv
	\sum_{kl\in B} 
	\tBraKet{k^B}{\hat{v}^{{\rm eff},A}}{l^B}
\end{equation}
%
with
%
\begin{equation}
 \hat{v}^{{\rm eff},A} 
	\equiv \sum_{ij\in A} {s} \left[ P^{AABB}_{ijkl} \hat{v}^{AA}_{ij} \right]  \;.
\end{equation}
%
Similarly for the overlap\hyp{}like term we have
%
\begin{equation}
 \sum_{ijkl} 
	{s} \left[ P^{AAAB}_{ijkl} \BraKet{AA}{AB} \right]
 \equiv
	\sum_{k\in A} \sum_{l\in B} 
	\tBraKet{k^A}{\hat{v}^{{\rm eff},A}}{l^B}
\end{equation}
%
and the associated effective potential reads
%
\begin{equation}
 \hat{v}^{{\rm eff},A} 
	\equiv \sum_{ij\in A} {s} \left[ P^{AAAB}_{ijkl} \hat{v}^{AA}_{ij} \right] \;.
\end{equation}
%
Note that the exchange\hyp{}like terms cannot be represented in terms of OEP's.
Gathering the above arguments, $\Delta^{AB}P$ adopts the following form
%
\begin{multline}
 \Delta^{AB}P \approx \Delta_{\rm ex}^{AB} P
	+ \sum_{i\in A} \sum_{j\in B} 
	  \tBraKet{i}{ \left\{ \hat{v}^{{\rm eff},AB}_{1(ij)} + \hat{v}^{{\rm eff},BA}_{1(ij)}  \right\} }{j}
	 \\ +
	\sum_{i\in A} \sum_{k\in A}
	  \tBraKet{i}{\hat{v}^{{\rm eff},BA}_{2(ik)} }{k}
	 +
        \sum_{j\in B} \sum_{l\in B}
          \tBraKet{j}{\hat{v}^{{\rm eff},AB}_{2(jl)} }{l}
\end{multline}
%
where the exchange\hyp{}like component is
%
\begin{equation}
	\Delta_{\rm ex}^{AB} P = 
	s \sum_{ik\in A} \sum_{jl\in B}
	P^{ABAB}_{ijkl} \BraKet{ij}{kl} \;,
\end{equation}
%
and
the overlap\hyp{}like and the Coulomb\hyp{}like 
OEP's are defined as follows
%
\begin{subequations}
\begin{align}
	\hat{v}^{{\rm eff},BA}_{1(ij)} &\equiv P_{ij}^{AB} \hat{h}
	+ s \sum_{mk\in A}  P^{AAAB}_{mkij} \hat{v}_{mk}^{AA} \;, \\
	%
	\hat{v}^{{\rm eff},BA}_{2(ik)} &\equiv P_{ik}^{AA} \hat{h} 
        + s \sum_{mk'\in A} P^{AAAA}_{ikmk'} \hat{v}_{mk'}^{AA}  \nonumber
        \\
	&+ \frac{1}{2}
        s \sum_{jl'\in B}  P^{AABB}_{ik'jl} \hat{v}_{jl'}^{BB} \;,
\end{align}
\end{subequations}
%
and accordingly for the twin operators $\hat{v}^{{\rm eff},AB}_{\text{$1$ and $2$}}$.

@@

An arbitrary one-electron potential of molecule $A$ acting on any state vector 
associated with molecule $A$ can be expanded in an auxiliary space centered 
on $A$ as
\begin{equation}
   \hat{v}^{A}\Ket{i} = \sum_{\xi\eta} \hat{v}^{A}\Ket{\xi} [{\bf S}^{-1}]_{\xi\eta} \BraKet{\eta}{i}
\end{equation}
under the necessary assumption that the auxiliary basis set is nearly complete,
i.e., 
$\sum_{\xi\eta} \Ket{\xi}[{\bf S}^{-1}]_{\xi\eta} \Bra{\eta} \cong 1$.
%In a special case when the basis set is orthogonal (e.g., molecular orbitals)
%the above relation simplifies to
%\begin{equation}
%   v\vert i) = \sum_{\xi} v\vert \xi) ( \xi \vert i)
%\end{equation}
The above general and exact expansion can be obtained by 
utilizing the resolution of identity or, equivalently, density fitting
in the complete space,
%
\begin{equation}
 \hat{v}^{A}\Ket{i} = \sum_{\xi} V^A_{i\xi} \Ket{\xi} \;.
\end{equation}
%
In the above equation,
the matrix ${\bf V}^A$
is the representation of the OEP operator $\hat{v}^{A}$
in the complete basis $\{ \varphi_\xi \}$.
By constructing the least\hyp{}squares objective function 
%
\begin{equation}
 Z[{\bf V}^A] = \int 
                     \left[    v({\bf r}_1) \phi_i({\bf r}_1) - \sum_\xi V^A_{i\xi} \varphi_\xi({\bf r}_1) \right]^2
d{\bf r}_1
\end{equation}
%
and requiring that
%
\begin{equation}
 \frac{\partial Z[{\bf V}^A]}{\partial V^A_{i\mu}} = 0 \text{ for all $\mu$}
\end{equation}
%
one finds the coefficients of the $i$th row of ${\bf V}^A$ to be
%
\begin{equation}
% {\bf v}_i = {\bf a}^{(i)} \cdot {\bf S}^{-1}
  V^A_{i\xi} = \sum_\eta [{\bf S}^{-1}]_{\xi\eta} a^{(i)}_\eta \;,
\end{equation}
%
where
%
\begin{subequations}
\begin{align}
 a^{(i)}_\eta &= \int \varphi^*_\eta({\bf r}) \hat{v}^A \phi_i({\bf r}) d{\bf r} \;,\\  
 S_{\eta\xi}  &= \int \varphi^*_\eta({\bf r}) \varphi_\xi({\bf r}) d{\bf r} \;.
\end{align}
\end{subequations}
%
%Since matrix elements of an OEP operator in auxiliary space can be computed 
%in the same way as the matrix elements with any other basis function, 
%one can formally write the following identity
%\begin{equation}
% (X \vert v\vert i) = \sum_{\xi\eta} S_{X \xi} [{\bf S}^{-1}]_{\xi\eta} (\eta\vert v\vert i)
%\end{equation}
%where $ X $ is an arbitrary orbital.
%When the other orbital
%does not belong to molecule $A$ but to the (changing) environment, it is 
%straightforward to compute the resulting matrix element, which is simply given as  
%\begin{equation}
%   (j_{\in B} \vert v^A \vert i_{\in A}) = \sum_\xi {S_{j\xi}} {G_{i\xi}}
%\end{equation}
%where $j$ denotes the other (environmental) basis function.
%
The working formula for $a^{(i)}_\eta$ can be found by applying 
potential for from Eq.~\eqref{e:v-eff}
along with the spectral representation of the effective density from Eq.~\eqref{e:d-spectral} 
which gives
%
\begin{equation}
 a^{(i)}_\eta = \sum_{x\in A} W_{\eta i}^{(x)} + 
 \sum_{\alpha\beta} D^A_{\alpha\beta} 
  \BraKet{\alpha\beta}{\eta i} \;.
%  \iint \frac{ \phi_\alpha^*({\bf r}_1)  \phi_\beta({\bf r}_1) \varphi^*_\eta({\bf r}_2) \phi_i({\bf r}_2)}
% {\vert {\bf r}_1 - {\bf r}_2 \vert }
% d{\bf r}_1 d{\bf r}_2
\end{equation}
%

@@

Density fitting scheme from previous section has practical disadvantage of a nearly-complete basis set
being usually very large (spanned by large amount of basis set vectors). Any non-complete basis set
won't work in the previous example. Since most of basis sets used in quantum chemistry do not form a nealy complete
set, it is beneficial to design a modified scheme in which it is possible to obtain the effective 
matrix elements of the OEP operator in a incomplete auxiliary space. This can be achieved by minimizing 
the following objective function
\begin{multline}
	Z[{\bf V}^A] = \iint 
        \Bigg\{
                     \left[    v^A({\bf r}_1) \phi_i({\bf r}_1) - \sum_\xi V^A_{i\xi} \varphi_\xi({\bf r}_1) \right]
                  \\ \times \frac{1}{\vert {\bf r}_1 - {\bf r}_2 \vert}  
                     \left[    v^A({\bf r}_2) \phi_i({\bf r}_2) - \sum_\xi V^A_{i\xi} \varphi_\eta({\bf r}_1) \right]
		    \Bigg\}
         d{\bf r}_1 d{\bf r}_2  \;.
\end{multline}
Requesting that 
\begin{equation}
 \frac{\partial Z[{\bf V}^A]}{\partial V^A_{i\mu}} = 0 \text{ for all $\mu$}
\end{equation}
one obtains
\begin{equation}
% {\bf G}^{(i)} = {\bf b}^{(i)} \cdot {\bf A}^{-1}
  V^A_{i\xi} = \sum_\eta [{\bf A}^{-1}]_{\xi\eta} b^{(i)}_\eta \;,
\end{equation}
where 
\begin{subequations}
\begin{align}
 b^{(i)}_\eta &= \iint 
                       \frac{ \varphi^*_\eta({\bf r}_1) \hat{v} \phi_i({\bf r}_2) } 
                            {\vert {\bf r}_1 - {\bf r}_2\vert}  
                 d{\bf r}_1 d{\bf r}_2 \;, \\
 A_{\eta\xi}  &= \iint 
                       \frac{ \varphi^*_\eta({\bf r}_1) \varphi_\xi({\bf r}_2) } 
                            {\vert {\bf r}_1 - {\bf r}_2\vert}  
                 d{\bf r}_1 d{\bf r}_2 \;.
\end{align}
\end{subequations}
%
%The symbol $ \vert\vert $ is to denote the operator $ r_{12}^{-1}$ and double integration over $ {\bf r}_1 $
%and $ {\bf r}_2 $. 
Note that, while $A_{\eta\xi}$ is a typical 2\hyp{}center ERI,
$b^{(i)}_\eta$ matrix elements are not at all trivial to evaluate
because the OEP operator, which contains integration over an electron coordinate,
is present inside the double integral. Therefore, the following triple integral
has to be computed
%
\begin{equation}
 b^{(i)}_\eta = \iiint 
           \frac{ \varphi^*_\eta({\bf r}_1) \phi_i({\bf r}_2)  \rho^{\rm eff}({\bf r}_3) }
            {\vert {\bf r}_1 - {\bf r}_2 \vert \vert {\bf r}_3 - {\bf r}_2 \vert}
           d{\bf r}_1 d{\bf r}_2 d{\bf r}_3
\end{equation}
%
Computation of all the necessery integrals of this kind is very costly 
and impractical even for medium sized molecules. 
However, one can introduce the effective potential in order to eliminate one integration. 
This can be achieved by performing additional density fitting in a nearly complete intermediate basis 
%
\begin{equation}
 \hat{v}^{\rm eff} \Ket{i} = \sum_\varepsilon^{\rm Int} H_{i\varepsilon} \Ket{\varepsilon} \;,
\end{equation}
%
where the symbol `Int' denotes the intermediate basis from a chosen resolution of identity.
The working equation is therefore
%
\begin{equation}
 b^{(i)}_\eta = \sum_\varepsilon^{\rm Int} H_{i\varepsilon} R_{\varepsilon\eta} 
\end{equation}
%
with
%
\begin{equation}
 R_{\varepsilon\eta} \equiv \iint 
                       \frac{ \varphi^*_\varepsilon({\bf r}_1) \varphi_\eta({\bf r}_2) } 
                            {\vert {\bf r}_1 - {\bf r}_2\vert}  
                 d{\bf r}_1 d{\bf r}_2 \;,
\end{equation}
%
which can be easily evaluated and the matrix elements of the OEP operator found.
%Therefore, in order to use this generalized density fitting scheme
%one must to compute two\hyp{}centre electron repulsion integrals
%as well as four\hyp{}centre asymmetric electron repulsion integrals of the type $ (\alpha\beta\gamma||\eta) $.
%The evaluation of such integrals is discussed in Appendix~\ref{a:mcmurchie-davidson}.
